#Project: Analysis of Steps Per Day.

For the first part of the analysis, I downloaded the file from the URL and load it in R usinf read.csv. No other preprocessing was done on this data set.The main dataset contains 3 variables - "steps", "date" and "interval"

```{r simulation,results="hide"}
library(ggplot2,quietly=TRUE)
library(data.table,quietly=TRUE)
library(dplyr,quietly=TRUE)
library(timeDate,quietly=TRUE)
url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url,"dataset.zip")
unzip("./dataset.zip")
f<-list.files()
data<-read.csv(f[1])
```

Once the data is loaded, I created a function "Q1" to obtain the mean total number of steps taken per day. The dataset is transformed using the aggregate function to calculate the total number of steps each day. NA values are ignored for thsi step. The function Q1 takes the entire dataset as the input and provides a listed output with a barplot of the average number of steps on each day, mean and median of the total number of steps each day. The function also provides the dataset.

```{r}
Q1<-function(data)
  {
  d<-aggregate(cbind(steps)~date,data,sum)
  steps_mean<-mean(d$steps)
  steps_med<-median(d$steps)
  print(paste0("The mean of the total steps per day is: ", round(steps_mean,digits=2)))
  print(paste0("The median of the total steps per day is: ", round(steps_med,digits=2)))
  g<-ggplot(d,aes(date))+geom_bar(aes(weight=steps))
  g<-g+theme(axis.text.x = element_text(angle = 90))
  print("The plot showing the total number of steps each day is as shown:")
  print(g)
  return (list(d,steps_mean,steps_med,g))
  }
A1<-Q1(data)
```

A second function Q2 is also created to calculate the average steps number of steps for each 5-minute interval. The function also takes in the raw data set as input and transforms it using an aggregate function to calculate a different dataset of the average number of steps in each 5-minute interval. The function provides a listed output of the transformed dataset, the 5-minute interval with the maximum number steps and a line plot showing the average steps for each 5-minute interval.

```{r}
Q2<-function(data)
  {
  d<-aggregate(cbind(steps)~interval,data,mean)
  max_avgspd<-max(d$steps)
  TI<-d[which.max(d$steps),"interval"]
  g<-ggplot(d,aes(interval,steps))+geom_line()
  print(paste0("The maximum average steps per day are: ",round(max_avgspd,digits=2)))
  print(paste0("And they occur at time intervals: ",TI))
  print("The plot showing the avg steps for each 5-minute time interval is as shown:")
  print(g)
  return(list(d,max_avgspd,g))
  }
A2<-Q2(data)
```

For imputing the missing valuves, I first, split the main data set into two parts:

- With NA values

- Without NA values

Using the nrow function, I calculated the total number of rows with NA values. Then I merged the average steps for each 5 minute interval (an output from the Q2 function) with the dataset with NA values. For this, I used the merge function and set "by" to the interval field. 

In the next step I removed the unnecessary column of the newly merged data set, using the select function (under dplyr package). Finaly, I combined the newly created data set and the original non-NA dataset using the rbind function and arranged it chronologically. I parsed this dataset through function Q1 to get the mean and median of the steps per day.


```{r}
d1<-data[is.na(data$steps),]
d2<-data[!(is.na(data$steps)),]
print(paste0("The number of NA values are: ",nrow(d1)))
d3<-merge(d1,as.data.frame(A2[1]),by="interval",all.x=is.na(d1$steps),suffixes=c(".x",""))
d3<-select(d3,steps,date,interval)
d3<-rbind(d3,d2)
d3<-arrange(d3,date,interval)
A3<-Q1(d3)
```

Finally, to compare using a panel plot, the average steps on weekdays and weekends, I first converted the date field into class "Date", using the as.Date function. Then, I added a factor column called "day", to identify using inbuilt functions in R, whether the date is a weekday or weekend. Using the facet grid functions of ggplot2, I created a one row, two column grid that shows two line plots of average steps on weekdays and weekends.

```{r}
d3$date<-as.Date(d3$date)
d3$day<-"Weekday"
d3[!isWeekday(d3$date),"day"]<-"Weekend"
d3$day<-as.factor(d3$day)
d<-aggregate(cbind(steps)~interval+day,d3,mean)
g<-ggplot(d,aes(interval,steps))+geom_line()
g<-g+facet_grid(.~day)
print("The panel plot showing the comparison of average number of steps on weekdays and weekends is as shown:")
print(g)
```

